---
title: "Eric_Hirsch_Assignment_3"
author: "Eric Hirsch"
date: "2/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-packages, message=FALSE}
library(openintro)
library(tinytex)
library(tidyverse)
library(dplyr)
library(stringr)
```


## Homework 3: Working with Strings

### Excercise 1. 

Using the 173 majors listed in fivethirtyeight.com’s College Majors dataset [https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/], provide code that identifies the majors that contain either "DATA" or "STATISTICS"

``` {r load data}
library(stringr)
library(dplyr)
dfMajors <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv", header= TRUE)
x <- filter(dfMajors, str_detect(dfMajors$Major, ("STATISTICS|DATA")))
x
        

```
## Excercise 2.

Write code that transforms the data below:

[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"

[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  

[9] "elderberry"   "lime"         "lychee"       "mulberry"    

[13] "olive"        "salal berry"

Into a format like this:

c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")

```{r fruits}

v1<-('"bell pepper"  "bilberry"     "blackberry"   "blood orange"')
v2<-('"blueberry"    "cantaloupe"   "chili pepper" "cloudberry"')
v3<-('"elderberry"   "lime"         "lychee"       "mulberry"')
v4<-('"olive"        "salal berry"')

vFruits<-str_c(v1, v2, v3, v4)
vFruits<-str_remove_all(vFruits, " {2,}")
vFruits<-str_c("c(",str_replace_all(vFruits, '\""|\" "', '\", "'),")")
writeLines(vFruits)

```
## Exercise 3.

Describe, in words, what these expressions will match:

a. (.)\1\1 - This will match sequences with \1\1 but it will read the slashes as escape characters. If properly escaped, this will match a 3 character sequence in which each character is the same.\
b. (.)(.)\\2\\1 - This will match a sequence which is a four character palindrome.\
c. (..)\1 - This will match sequences with \1 after two characters but it will read the slash as an escape character.  If properly escaped, this will match a four character sequence in which a pair of characters is repeated.\
d. (.).\\1.\\1 - This will match a five character sequence in which the first, third and fifth character are the same.\
e. (.)(.)(.).*\\3\\2\\1 - This will match a character sequence of 6 or more in which the last three characters are a reverse of the first three.

```{r}
words <- c("abcdefbca", "remembmerabba", "eleven\1\1", "church", "Missojh;ougovssippi")
x<-stringr::str_view(words, "(.)(.)(.).*\\3\\2\\1", match = TRUE)
x
```

## Exercise 4.

Construct regular expressions to match words that:

a. Start and end with the same character.
    ^(.).*\\1$

b. Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.) 
    (.)(.).*\\1\\2

c. Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)
    (.).*\\1.*\\1

```{r 2}
words <- c("abcdefba", "remebmerabbar", "elevn\1\1", "churcch", "Missojh;ougovsippi")
x<-stringr::str_view(words, "(.).*\\1.*\\1", match = TRUE)
x
```

