---
title: "Poverty and Diabetes: Is There A Connection?"
Author: "Eric Hirsch"
output: 
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning=FALSE)
```
```{r ech=TRUE, message = FALSE, warning=FALSE}
library(ROSE)
library(tidyverse)
library(magrittr)
library(janitor)
library(tidyverse)
library(RODBC)
library(pscl)
library(caret)
library(car)
library(InformationValue)
library(rvest)
library(jsonlite)
library(xml2)
library(XML)
library(stringr)
library(kableExtra)
```
## Diabetes and Poverty: Is There a Connection?

To investigate this question We will use data from 4 different sources: 

1. Health data from the New York Department of Health (in json format accessed through APIs)
2. Zip code level data on population size and poverty (scraped from html tables on the web)
3. Files from a proprietary survey of economic needs, stripped of identity markers (stoed in a csv file)
4. Tables of client health information in a proprietary SQL Server database, stripped of identity markers and password protected through keyring (SQL Server database on Azure)

After collecting the data we will store it in the database so that we can easily combine the datasets in productive ways using SQL.  In particular, we will generate dataframes that contain zip code level information on poverty and public health, and client level information on economic needs and individual client health.  Then we will test for correlations and other relationships within the data.

Note - Certain operations that load data and are very time consuming will be contained within functions that can be run or not by changing the value of the boolean variable "run" which appears at the top of each function.

## Loading The Data

### 1. New York DOH Data

We begin by reading the New York DOH data from the APIs.  The RSocrata library makes this easy.
```{r}
library("RSocrata")

#This operation is time consuming.  Set run to TRUE the first time you run ii, and FALSE afterward.
run=FALSE 

if(run) {
#Community Health: Diabetes mortality rate per 100,000: 2009 - 2011
dfMortality2011 <- read.socrata(
  "https://health.data.ny.gov/resource/u98s-c3hg.json")

#U.S. Chronic Disease Indicators: Diabetes
dfDiabetesIndicators <- read.socrata(
  "https://chronicdata.cdc.gov/resource/f8ti-h92k.json")

#Community Health Obesity and Diabetes Related Indicators: 2008 - 2012
dfCommunityHealthObesityAndDiabetes <- read.socrata(
  "https://health.data.ny.gov/resource/tchg-ruva.json")

#Community Health: Age-adjusted percentage of adults with physician diagnosed diabetes: 2008 - 2009
dfAdultsDiagnosed <- read.socrata(
  "https://health.data.ny.gov/resource/9j5w-7zpd.json")

#AH Provisional Diabetes Death Counts, 2020
dfDeath <- read.socrata(
  "https://data.cdc.gov/resource/qdcb-uzft.json")

#500 Cities: Diagnosed diabetes among adults aged >=18 years
dfCityComparison <- read.socrata(
  "https://chronicdata.cdc.gov/resource/cn78-b9bj.json")

#Conditions contributing to deaths involving coronavirus disease 2019 (COVID-19), by age group and state, United States.
dfDeathsForCovid <- read.socrata(
  "https://data.cdc.gov/resource/hk9y-quqm.json")

#Community Health: Diabetes Short-term Complications Hospitalization Rate per 10,000 - Aged 18+ Years by County Map: Latest Data
dfHospitalizations <- read.socrata(
  "https://health.data.ny.gov/resource/xuwq-ppg8.json")

#Medicaid Chronic Conditions, Inpatient Admissions and Emergency Room Visits by Zip Code: Beginning 2012
dfMedicaidByZip <- read.socrata(
  "https://health.data.ny.gov/resource/2yck-xisk.json")

}

```
```{r}
#This operation is time consuming.  Set run to TRUE the first time you run ii, and FALSE afterward.
run=FALSE 

if(run) {
#Medicaid Program Enrollment by Month: Beginning 2009
dfMedicaidEnrollment <- read.socrata(
  "https://health.data.ny.gov/resource/m4hz-kzn3.json")
}
```


### 2. Zip Code Level Population and Poverty Data Scraped From the web

We scrape and then clean 17 pages of zipatlas data to compile a table of zip codes, population levels and poverty levels.

```{r html}
library(XML)

#This operation is time consuming.  Set run to TRUE the first time you run ii, and FALSE afterward.
run=FALSE

if(run) {


x="http://www.zipatlas.com/us/ny/zip-code-comparison/population-below-poverty-level.htm"
dfPopAndPovertyLevel = as.data.frame(readHTMLTable(x, header=T,which=5,strings2factors=F))

for (i in 1:16) {
x= str_c("http://www.zipatlas.com/us/ny/zip-code-comparison/population-below-poverty-level.", i, ".htm")
dfx = as.data.frame(readHTMLTable(x, header=T,which=5,strings2factors=F))
dfPopAndPovertyLevel=rbind(dfPopAndPovertyLevel, dfx)
}
}

#This operation is time consuming.  Set run to TRUE the first time you run ii, and FALSE afterward.
run=FALSE

if(run) {
  
dfPopAndPovertyLevel %<>%
  filter(str_length(V2)==5 & !is.na(V4)) %<>%
  rename(number = V1, zip=V2, location=V3, city=V4, population=V5, percent_below_poverty=V6, rank=V7) %<>%
  mutate(population = as.numeric(gsub(",", "", population))) %<>%
  mutate(percent_below_poverty = as.numeric(gsub(" %", "", percent_below_poverty))) %<>%
  mutate(rank = as.numeric(gsub("#", "", rank)))

}

```


### 3. CSV file of a Survey of Client Economic Needs

This file is the result of a client survey taken at my workplace of over 2,000 clients.  We did not create the survey instrument - the file was provided by the organization that did.  The Ids have been mathced to those of the clients in the SQL database.  The file is otherwise free of any personal identifying information.  

```{r}

#This operation is time consuming.  Set run to TRUE the first time you run ii, and FALSE afterward.
run=FALSE

if(run) {
dfSDOH_raw <- read.csv("D:\\Buffer\\SDOH__En__1_0 (3).csv", encoding = "UTF-8")
}

run=FALSE

if(run) {
dfBroadband_raw <- read.csv("D:\\RStudio\\CUNY_607\\Projects\\Final Project\\Broadband_Adoption.csv", encoding = "UTF-8")
}

dfListOfNeeds <- read.csv("D:\\RStudio\\CUNY_607\\Projects\\Final Project\\ListOfNeeds.csv", encoding = "UTF-8")
dfListOfNeeds %<>%
  rename(Need=X.U.FEFF.Need)

```

The survey asks these questions:

```{r}

kableExtra::kable(dfListOfNeeds)


```
\

Since the survey instrument and resultant data were created and delivered by an outside agency, the data needs extensive reforming. The file contains 800,000 observations with only two columns- the columns are "Field" and "Value".  Each record takes up 211 observations.

We begin by pivoting wider every 211 records, taking our names from "Field" and our values from "Value", and binding each to a final dataframe.

```{r}

run=FALSE

if (run) {
dfSDOH <- dfSDOH_raw %>%
    filter(as.numeric(Num) <=211) %>%
    pivot_wider(Org, names_from = "Field", values_from = "Value")

dfFinal <- dfSDOH

x=211

for (i in 1: 3752)
{

dfSDOH <- dfSDOH_raw %>%
    filter(as.numeric(Num) > x & as.numeric(Num)<= (x+211)) %>%
    pivot_wider(Org, names_from = "Field", values_from = "Value")

x=x+211

dfFinal <- rbind(dfFinal, dfSDOH)
}
}
```

We eliminate bad records that have no ClientID.

```{r}
#This operation is time consuming.  Set run to TRUE the first time you run ii, and FALSE afterward.
run=FALSE

if(run) {
dfFinal %<>%
  filter (ClientID > 0)
}

```

We select the columns which describe needs, and rename them to reflect the type of need.

```{r}

NeedsVector = c("Need_Food",  "Need_SafePlace",  "Need_LoseHousing",  "Need_Job",  "Need_AffordNeeds",  "Need_HighSchool",  "Need_HelpUnderstanding",  "Need_Sad",  "Need_Childcare",  "Need_Clothes",  "Need_Transport",  "Need_Safe") 


#This operation is time consuming.  Set run to TRUE the first time you run ii, and FALSE afterward.
run=FALSE

if(run) {

dfFinal2 <- dfFinal %>%
  rename(Need_Food=DoYouOrYou, Need_SafePlace = DoYouHaveA, Need_LoseHousing = AreYouWorr, Need_Job = DoYouHaveA1, Need_AffordNeeds = AreYouAble, Need_HighSchool = DoYouHaveA2, Need_HelpUnderstanding=DoYouEverN, Need_Sad = SocialAreY, Need_Childcare = ChildCareD, Need_Clothes = DoYouEverH, Need_Transport = DoYouHaveA3, Need_Safe = DoYouFeelS )

dfFinal3 <- dfFinal2 %>%
  select(ClientID, starts_with('Need_'))
}

```

Each of the 12 need columns is binary, containing either a yes or no in response to a question about a need. However, sometimes 'yes' means the client has a need ("Do you or your family ever go hungry") and sometimes 'yes' means the client does not have a need ("Do you feel safe physically and emotionally").

We will recode the questions so that 1 means 'has a need' and 0 means 'does not have a need' for all variables.

```{r}

#This operation is time consuming.  Set run to TRUE the first time you run ii, and FALSE afterward.
run=FALSE

if(run) {
QCorrect<- function(x){
    ifelse(x=="Yes", 1,0)
}

QReverse<- function(x){
    ifelse(x=="No", 1,0)
}
 
dfFinal4 <- dfFinal3 %>%
  select(Need_Food, Need_Sad, Need_HelpUnderstanding, Need_Childcare, Need_Clothes, Need_LoseHousing) %>%
   mutate_all(QCorrect)

dfFinal5 <- dfFinal3 %>%
  select(Need_SafePlace, Need_Job, Need_AffordNeeds, Need_HighSchool, Need_Transport, Need_Safe) %>%
   mutate_all(QReverse)

dfFinal6 <- dfFinal3 %>%
  select(ClientID)

dfFinal6$ClientID <- as.numeric(as.character(dfFinal6$ClientID))

dfFinal10 <- cbind(dfFinal6, dfFinal4, dfFinal5)
}

```


### 4. SQL Server Database Data

I built this database for my workplace.  It contains information on client health screens for diabetes, hypertension and other chronic diseases.  The database contains 2 tables: 

a. A table of 5,433 unique clients with demographic data and the client's first A1C result (A1C is a test that indicates diabetes.)

b. A table of 39,491 observations for 8,546 unique clients with A1C, Blood Pressure and other health screen data.

The data is password protected by keyring.  Here we create a connection.

```{r keyring-set-up, echo=FALSE}
#install.packages("keyring")
library(keyring)
#keyring_create("EH_606_keyring")
keyring_unlock("EH_606_keyring")
#key_set(service = "EH_606_username", keyring = "EH_606_keyring")
#key_set(service = "EH_606_pwd", keyring = "EH_606_keyring")
```

```{r db connect, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# This is the connection string:

strConnection <- paste0(
  'Driver={ODBC Driver 13 for SQL Server};
   Server=tcp:ehtmp.database.windows.net,1433;
   Database=HC_A1C;
   Encrypt=yes;
   TrustServerCertificate=no;
   Connection Timeout=30;',
   'Uid=',keyring::key_get(service = "EH_606_username", keyring = "EH_606_keyring"),';',
   'Pwd=', keyring::key_get(service = "EH_606_pwd", keyring = "EH_606_keyring"), ';'
)

library(RODBC)
dbConnection <- odbcDriverConnect(strConnection)
#keyring_lock
```

## Loading all of the Datasets into the SQL Server Database

If we load our datasets into persistent storage (the SQL Sever database), then we can use SQL to easily combine data from several tables and read the result back into a dataframe. Just as important, many of the datasets in this project are very large and take a long time to load.  Therefore we can filter and prepare them, read them into the database, and then use the database exclusively to read them back to cut down on load time. 

We begin with 3 NYDOH files: MedicaidByZip, a NYDOH file with over a million records on medicaid information; dfAdultsDiagnosed which has 63 records about Diabetes diagnoses throughout New York State; and dfCommunityHealthObesityAndDiabtes, a dataset with 2733 records with detailed county-level information about diabetes.

```{r}
#This operation is time consuming.  Set run to TRUE the first time you run ii, and FALSE afterward.
run=FALSE
if(run) {

DropTable <- sqlQuery(dbConnection, "DROP TABLE IF EXISTS tblMedicaidByZip")
PopulateTable <- sqlSave(dbConnection, dfMedicaidByZip, "tblMedicaidByZip", append=TRUE)
}

```

```{r}
#This operation is time consuming.  Set run to TRUE the first time you run ii, and FALSE afterward.
run=FALSE
if(run) {

DropTable <- sqlQuery(dbConnection, "DROP TABLE IF EXISTS tblAdultsDiagnosed")
PopulateTable <- sqlSave(dbConnection, dfAdultsDiagnosed, "tblAdultsDiagnosed", append=TRUE)
}

```

```{r}
#This operation is time consuming.  Set run to TRUE the first time you run ii, and FALSE afterward.
run=FALSE
if(run) {

DropTable <- sqlQuery(dbConnection, "DROP TABLE IF EXISTS tblCommunityHealth")
PopulateTable <- sqlSave(dbConnection, dfCommunityHealthObesityAndDiabetes, "tblCommunityHealth", append=TRUE)
}

```

Next we clean and load the scraped dataset with 1711 zip code level observations on population and poverty in NY State:

```{r}

#This operation is time consuming.  Set run to TRUE the first time you run ii, and FALSE afterward.
run=FALSE

if(run) {
DropTable <- sqlQuery(dbConnection, "DROP TABLE IF EXISTS tblPopAndPovertyLevel")
PopulateTable <- sqlSave(dbConnection, dfPopAndPovertyLevel, "tblPopAndPovertyLevel", append=TRUE)
}

```

Next we load the survey data into a table called tblNeeds:

```{r}

#This operation is time consuming.  Set run to TRUE the first time you run ii, and FALSE afterward.
run=FALSE

if(run) {

  
DropTable <- sqlQuery(dbConnection, "DROP TABLE IF EXISTS tblNeeds")
PopulateTable <- sqlSave(dbConnection, dfFinal10, "tblNeeds", append=TRUE)
}

```

Finally, we add the broadband_by_zip_code dataframe

```{r}
#This operation is time consuming.  Set run to TRUE the first time you run ii, and FALSE afterward.
run=FALSE

if(run) {

DropTable <- sqlQuery(dbConnection, "DROP TABLE IF EXISTS tblBroadBand")
PopulateTable <- sqlSave(dbConnection, dfBroadband_raw, "tblBroadband", append=TRUE)
}

```

## Creating Dataframes for Analysis By Joining Tables in the SQL Server Database

We will create the following dataframes by combining elements from a number of different tables in the database:

1. dfZipInfo, a dataframe with 99 observations at the zip code level with information on population level, poverty level, medicaid Info, average A1C from client data, and broadband info.

2. dfdNeedsInfo_Diabetes and dfNeedsInfo_Hypertension, dataframe with client-level observations with information on needs from the survey, chronic diseases, demographics and initial A1C. The diabetes database contains 348 observations and only includes only those clients who took the survey and have been screened for diabetes.The hypertension database contains 463 observations and only includes only those clients who took the survey and have been screened for hypertension.  

3. dfA1CAndDemographics, a dataframe with 5433 observations at the client level with information on demographics and initial A1C readings.

```{r}

dfZipInfo_raw <- sqlQuery(dbConnection, "SELECT tblMedicaidByZip.zip_code, tblMedicaidByZip.major_diagnostic_category, tblPopAndPovertyLevel.population, tblPopAndPovertyLevel.percent_below_poverty, tblMedicaidByZip.ip_admits, tblMedicaidByZip.er_visits, tblA1CAndDemographics.ClientID, tblA1CAndDemographics.Zip, tblA1CAndDemographics.InitialA1C, tblA1CAndDemographics.Gender, tblA1CAndDemographics.RaceandEthnicity, tblA1CAndDemographics.Age, tblBroadband.NoInternetAccessPercentageOfHouseholds
FROM (tblMedicaidByZip INNER JOIN (tblA1CAndDemographics INNER JOIN tblPopAndPovertyLevel ON tblA1CAndDemographics.Zip = tblPopAndPovertyLevel.zip) ON tblMedicaidByZip.zip_code = tblPopAndPovertyLevel.zip) INNER JOIN tblBroadband ON tblPopAndPovertyLevel.zip = tblBroadband.ZipCode;
")


dfZipInfo10 <- dfZipInfo_raw %>%
  filter(major_diagnostic_category=="Diabetes Mellitus") %>%
  na.omit() %>%
  group_by(zip_code) %>%
  summarize(n=n(), ave_A1C=mean(InitialA1C), pov=mean(percent_below_poverty), pop=mean(population), NoInternet=mean(NoInternetAccessPercentageOfHouseholds*100), percent_ip = sum(ip_admits/population), percent_er=sum(er_visits/population))


dfZipInfo_SI <- dfZipInfo10 %>%
  filter(as.numeric(zip_code) >= 10301 & as.numeric(zip_code) <= 10314) 
```

```{r}

dfA1CAndDemographics <- sqlQuery(dbConnection, "SELECT * FROM tblA1CAndDemographics")

dfA1CAndDemographics %<>% 
  mutate(BlackOrNot = ifelse(RaceandEthnicity=="Black/African American", 1, 0)) %<>%
  mutate(MaleOrNot = ifelse(Gender=="Male", 1, 0)) %>%
  mutate(A1CAboveMean = ifelse(InitialA1C > 7, 1, 0))

```

```{r}

dfNeedsInfo_raw <- sqlQuery(dbConnection, "SELECT tblNeeds.ClientID, tblChronic.Chronic, tblReadings.Type,  tblNeeds.Need_Food, tblNeeds.Need_Sad, tblNeeds.Need_HelpUnderstanding, tblNeeds.Need_Childcare, tblNeeds.Need_Clothes, tblNeeds.Need_LoseHousing, tblNeeds.Need_SafePlace, tblNeeds.Need_Job, tblNeeds.Need_AffordNeeds, tblNeeds.Need_HighSchool, tblNeeds.Need_Transport, tblNeeds.Need_Safe
FROM tblReadings INNER JOIN (tblChronic RIGHT JOIN tblNeeds ON tblChronic.ClientID = tblNeeds.ClientID) ON tblReadings.ClientID = tblNeeds.ClientID
WHERE tblReadings.Type='A1C';")

    
dfNeedsInfo <- dfNeedsInfo_raw %>%
    group_by(ClientID) %>%
  
    mutate(DiabetesOrNot = ifelse(Chronic == "Diabetes Type I"|Chronic=="Diabetes Type II"|Chronic == "Diabetes (doctor denied)"|Chronic == "Diabetes type I", 1, 0)) %>%
    mutate(HypertensionOrNot = ifelse(Chronic == "Hypertension", 1,0)) %>%
    mutate(HypertensionOrDiabetesOrNot = ifelse(Chronic == "Diabetes Type I"|Chronic=="Diabetes Type II"|Chronic == "Hypertension", 1,0)) %>%
  
    summarize(sum(DiabetesOrNot), sum(HypertensionOrNot), sum(HypertensionOrDiabetesOrNot), across(NeedsVector, mean, na.rm = TRUE)) %>%
  
    rename(DiabetesOrNot="sum(DiabetesOrNot)") %>%
      rename(HypertensionOrNot="sum(HypertensionOrNot)") %>%
      rename(HypertensionOrDiabetesOrNot="sum(HypertensionOrDiabetesOrNot)") %>%
  
      mutate(DiabetesOrNot = ifelse(DiabetesOrNot>0, 1, 0)) %>%
      mutate(HypertensionOrNot = ifelse(HypertensionOrNot>0, 1, 0)) %>%
      mutate(HypertensionOrDiabetesOrNot = ifelse(HypertensionOrDiabetesOrNot>0, 1, 0)) %>%
      mutate(NumOfNeeds = Need_Sad+Need_HelpUnderstanding+Need_Childcare+Need_Clothes+ Need_LoseHousing+Need_SafePlace+ Need_Job+Need_AffordNeeds+Need_HighSchool+Need_Transport+Need_Safe) 

dfNeedsInfo$DiabetesOrNot <- replace(dfNeedsInfo$DiabetesOrNot, is.na(dfNeedsInfo$DiabetesOrNot), 0)


```

```{r}

dfNeedsInfo2_raw <- sqlQuery(dbConnection, "SELECT tblNeeds.ClientID, tblChronic.Chronic, tblReadings.Type,  tblNeeds.Need_Food, tblNeeds.Need_Sad, tblNeeds.Need_HelpUnderstanding, tblNeeds.Need_Childcare, tblNeeds.Need_Clothes, tblNeeds.Need_LoseHousing, tblNeeds.Need_SafePlace, tblNeeds.Need_Job, tblNeeds.Need_AffordNeeds, tblNeeds.Need_HighSchool, tblNeeds.Need_Transport, tblNeeds.Need_Safe
FROM tblReadings INNER JOIN (tblChronic RIGHT JOIN tblNeeds ON tblChronic.ClientID = tblNeeds.ClientID) ON tblReadings.ClientID = tblNeeds.ClientID
WHERE tblReadings.Type='Blood Pressure';")

    
dfNeedsInfo2 <- dfNeedsInfo2_raw %>%
    group_by(ClientID) %>%
  
    mutate(DiabetesOrNot = ifelse(Chronic == "Diabetes Type I"|Chronic=="Diabetes Type II"|Chronic == "Diabetes (doctor denied)"|Chronic == "Diabetes type I", 1, 0)) %>%
    mutate(HypertensionOrNot = ifelse(Chronic == "Hypertension", 1,0)) %>%
    mutate(HypertensionOrDiabetesOrNot = ifelse(Chronic == "Diabetes Type I"|Chronic=="Diabetes Type II"|Chronic == "Hypertension", 1,0)) %>%
  
    summarize(sum(DiabetesOrNot), sum(HypertensionOrNot), sum(HypertensionOrDiabetesOrNot), across(NeedsVector, mean, na.rm = TRUE)) %>%
  
    rename(DiabetesOrNot="sum(DiabetesOrNot)") %>%
      rename(HypertensionOrNot="sum(HypertensionOrNot)") %>%
      rename(HypertensionOrDiabetesOrNot="sum(HypertensionOrDiabetesOrNot)") %>%
  
      mutate(DiabetesOrNot = ifelse(DiabetesOrNot>0, 1, 0)) %>%
      mutate(HypertensionOrNot = ifelse(HypertensionOrNot>0, 1, 0)) %>%
      mutate(HypertensionOrDiabetesOrNot = ifelse(HypertensionOrDiabetesOrNot>0, 1, 0)) %>%
      mutate(NumOfNeeds = Need_Sad+Need_HelpUnderstanding+Need_Childcare+Need_Clothes+ Need_LoseHousing+Need_SafePlace+ Need_Job+Need_AffordNeeds+Need_HighSchool+Need_Transport+Need_Safe) 

dfNeedsInfo2$HypertensionOrNot <- replace(dfNeedsInfo2$HypertensionOrNot, is.na(dfNeedsInfo2$HypertensionOrNot), 0)


```

These are the final dataframes which will be used for analysis:

1. dfZipIno

```{r}

kable(psych::describe(dfZipInfo10), "html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
```

2. dfNeedsInfo_Diabetes

```{r}
kable(psych::describe(dfNeedsInfo), "html") %>% kable_styling("striped") %>% scroll_box(width = "100%")

```

3. dfNeedsInfo_Hypertension

```{r}
kableExtra::kable(psych::describe(dfNeedsInfo2), "html") %>% kable_styling("striped") %>% scroll_box(width = "100%")

```

4. dfA1CAndDemographics

```{r}

kableExtra::kable(psych::describe(dfA1CAndDemographics), "html") %>% kable_styling("striped") %>% scroll_box(width = "100%")

```

## Analysis - Exploring the Connection Between Poverty and Diabetes


The link between diabetes and poverty can be difficult to get at directly, as many poorer people who have diabetes are not aware of it, and diabetes incidence is often not available in such a way it can be correlated poverty (for example, it may only be available at the county level.)  The purpose of this analysis is to get at the correlation through other means: hospitalization rates, looking at a cohort of clients who have all been tested for diabetes, and A1C levels.

### Poverty and diabetes at the county and zip code level

Richmond county (Staten Island) is a relatively middle class borough.  The NY DOH reports that the incidence of diabetes in Staten Island is 8.5%.  The chart below shows the incidence of diabetes in New York City. Staten Island is the second lowest.  

```{r}


dfIncidenceOfDiabetes <- sqlQuery(dbConnection, "SELECT * FROM tblAdultsDiagnosed Where region_name = 'New York City';")


ggplot(dfIncidenceOfDiabetes, aes(percentage_rate, reorder(county_name, as.numeric(percentage_rate)))) +
  geom_col() +
  ggtitle("Incidence of Diabetes by New York City County")

```

However, among the 292 clients from our survey who were also screened for diabetes, the incidence is much higher at 19%.  What is happening here?

```{r}

  diab_percent <- dfNeedsInfo %>%
      summarize(PercentDiabetes = sum(DiabetesOrNot/n()))
  diab_percent
```

While Richmond county is middle class, the clients in our survey are economically disadvantaged - all are from poorer zipcodes seeking assistance from a community health organization.

However, consider the hospitalization rate per 100,000 for all individuals with diabetes in Staten Island:


```{r}
dfCommunityHealth <- sqlQuery(dbConnection, "SELECT * from tblCommunityHealth;")

HospRate <- dfCommunityHealth %>%
  filter(indicator=="Diabetes hospitalization rate per 10,000 (any diagnosis)" & county_name=="Richmond") %>%
  mutate(percentage_rate = percentage_rate/10) %>%
  select(percentage_rate)
HospRate

```

Now consider these scatterplots of hospital admissions for diabetes patients with Medicaid, against indicators of poverty per zip code.  

```{r}

a <- ggplot(dfZipInfo_SI, aes(pov, percent_ip)) +
  geom_point() +
  ggtitle("Poverty And Medicaid Hospitalizations") +
   theme(plot.title = element_text(size = 10, face="bold")) +
      xlab("Percent below the poverty line") +
  ylab("Hospitalizations per 100,000") +
  geom_hline(yintercept=26.4, size=2, color='red', linetype='dotted')

b <- ggplot(dfZipInfo_SI, aes(NoInternet, percent_ip)) +
  geom_point() +
  ggtitle("Poverty And Medicaid Hospitalizations") +
   theme(plot.title = element_text(size = 10, face="bold")) +
      xlab("Percent without internet") +
  ylab("Hospitalizations per 100,000") +
  geom_hline(yintercept=26.4, size=2, color='red', linetype='dotted')


gridExtra::grid.arrange(a,b,ncol=2)

```

While the strong relationship between poverty level and hospitalization rate for medicaid recipients is likely due in part to the fact that there are more medicaid recipients in poorer zip codes, nonetheless we can see from these graphs that the 5 poorest zip codes have higher rates of hospitalization for medicaid resipients alone than the average hospitalization rate for all recipients.

### Poverty and diabetes at the individual level

The second graph suggests that poverty affects individuals in a direct way - lack of internet may lead to lack of knowledge and poor health choices.

Our survey data allow us to examine diabetes and poverty at the individual level. We will look first to see if incidence of diabetes and particular economic and social needs are correlated.

We examine the correlation between the incidence of diabetes and social needs through chi squares and a multiple regression:

```{r, warning=FALSE}

m10 <- glm(DiabetesOrNot ~ Need_Sad + Need_HelpUnderstanding + Need_Childcare + Need_Clothes + Need_LoseHousing + Need_SafePlace + Need_Job + Need_AffordNeeds + Need_HighSchool + Need_Transport + Need_Safe, family = "binomial", data=dfNeedsInfo)
summary(m10)

dfNeedsChi <- dfNeedsInfo %>%
  select(DiabetesOrNot, NeedsVector) %>%
  summarise_all(funs(chisq.test(.,dfNeedsInfo$DiabetesOrNot)$p.value))
kableExtra::kable(head(dfNeedsChi), "html") %>% kable_styling("striped") %>% scroll_box(width = "100%")

```

Those clients who cannot afford basic needs have a higher incidence of diabetes in both the chi squares and regression analysis.  However, we cannot reject the null hypothesis for other needs.

We can also investigate whether we find the same pattern for hypertension clients.

```{r}

m10 <- glm(HypertensionOrNot ~ Need_Sad + Need_HelpUnderstanding + Need_Clothes + Need_LoseHousing + Need_SafePlace + Need_Job + Need_AffordNeeds + Need_HighSchool + Need_Transport + Need_Safe, family = "binomial", data=dfNeedsInfo2)
summary(m10)

dfNeedsChi <- dfNeedsInfo2 %>%
  select(HypertensionOrNot, NeedsVector) %>%
  summarise_all(funs(chisq.test(.,dfNeedsInfo2$HypertensionOrNot)$p.value))
kableExtra::kable(head(dfNeedsChi), "html") %>% kable_styling("striped") %>% scroll_box(width = "100%")



```

In this case the consistent factor is whether a client does not have clothing for their family (this factor appears as significant in both the chi square and the regression).  However, the finding is surprising as the coeficient is negative, in other words, clothing need is negatively correlated with high blood pressure.  

we investigate these clients further to see if any confounding variable is obvious:

```{r}
dfJoin <- dfNeedsInfo2 %>% 
  inner_join(dfA1CAndDemographics, by="ClientID") %>%
    filter(Need_Clothes==1)
dfJoin
```

What stands out most is that the other clients mostly don't have diabetes diagnoses.  Since most clients are screened for both, this would appear to be a non-random cohort who were approached or surveyed differently.  More investigation is needed.

3. predictive model

No help

data.rose <- ROSE::ROSE(DiabetesOrNot ~ Need_AffordNeeds, data=train, seed=1)$data


model <- glm(DiabetesOrNot ~ Need_AffordNeeds, family="binomial", data=data.rose)

table(data.rose$DiabetesOrNot)

table(data.rose$HypertensionOrDiabetesOrNot)

```{r}

#https://www.statology.org/logistic-regression-in-r/
#make this example reproducible
set.seed(1)

#Use 70% of dfNeedsInfoset as training set and remaining 30% as testing set
sample <- sample(c(TRUE, FALSE), nrow(dfNeedsInfo2), replace=TRUE, prob=c(0.7,0.3))
train <- dfNeedsInfo2[sample, ]
test <- dfNeedsInfo2[!sample, ]  

#fit logistic regression model
model1 <- glm(HypertensionOrNot ~ Need_HighSchool + Need_Clothes + Need_Sad, family="binomial", data=train)

summary(model1)

#model <- glm(HypertensionOrDiabetesOrNot ~ NumOfNeeds, family="binomial", data=train)
#model
#disable scientific notation for model summary
#options(scipen=999)


#Shows fit - ranges from 0 to almost 1
pscl::pR2(model1)["McFadden"]

#Ranks importance
caret::varImp(model1)

#multicolinearity (VIF)
car::vif(model1)

#calculate probability of default for each individual in test dataset
predicted <- predict(model1, test, type="response")
p2 <- as.data.frame(predicted)

p3 <- test %>%
  select(HypertensionOrNot)

Predictions <- cbind(p2, p3)
Predictions$predicted <- ifelse(Predictions$predicted >.405, 1, 0)
Predictions$predicted=as.factor(Predictions$predicted)
Predictions$HypertensionOrNot=as.factor(Predictions$HypertensionOrNot)

psych::describe(Predictions)

#find optimal cutoff probability to use to maximize accuracy
optimal <- InformationValue::optimalCutoff(test$HypertensionOrNot, predicted)[1]
optimal

#ConfusionMatrix doesn't work
caret::confusionMatrix(Predictions$HypertensionOrNot, Predictions$predicted)

#calculate sensitivity
#sensitivity(test$HypertensionOrDiabetesOrNot, predicted)

#calculate specificity
#specificity(test$HypertensionOrDiabetesOrNot, predicted)

#calculate total misclassification error rate
#misClassError(Predictions$HypertensionOrDiabetesOrNot, Predictions$predicted, threshold=optimal)


```

Lastly we look at race and other demographics.

```{r}


psych::describe(dfA1CAndDemographics)

rBase1 <- lm(InitialA1C ~ BlackOrNot + MaleOrNot + Age,  data = dfA1CAndDemographics)
summary(rBase1)

```

Here we can see that race, gender and age all affect the initial A1C at a client's first screening.

### Conclusion

This study touched the surface of the question, "how are poverty and diabetes connected?'  We found statistical associations between poverty and diabetes hospitalization rates, between being unable to afford basic needs and diabetes incidence rates, and a connection between high A1C (indicative of diabetes) and race, age and gender.  We also found evidnece to suggest our study has nonradom elements.  Clearly there is more to explore here.