---
title: "Eric_Hirsch_607_DataScienceInContext"
author: "Eric Hirsch"
date: "`r Sys.Date()`"
output: 
  tufte::tufte_html: default
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
```

```{r load-packages, message=FALSE}
library(openintro)
library(tinytex)
library(tidyverse)
library(stringr)
library(magrittr)
library(gridExtra)
library(tufte)
library(knitr)
library(ggthemes)
library(extrafont)
#font_import()
loadfonts()
```

## 
## <span style="color:darkorchid"><font size="6">__Principle 1: Style expresses substance__</font size></span>

Elegance and style are not superfluous to or, even worse, antithetical to good data design - they are central to readability and clarity.

Consider the elegance of this table:    

``` {r load data}
dfFlights_raw <- as.data.frame(read.delim("https://raw.githubusercontent.com/ericonsi/CUNY_607/main/Projects/Project%202/Flights.csv", header = TRUE, stringsAsFactors = TRUE, sep=","))
```

```{marginfigure, echo=TRUE}

<span style="color:darkorchid"><font size="6">__Principle 2. Form expresses function__ </font size></span>

```

```{marginfigure, echo=TRUE}  
Placement, color, width - all express meaning If it's a side comment, put it on the side.  For example, this would be a good place to make a side comment about the table to the left.

```

```{r bvkf}
dfFlights <- dfFlights_raw %>% 
  drop_na(Phoenix) %>%
  rename(c(Airline =  "X", Status = "X.1")) %>%
  mutate_all(list(~na_if(.,"")))

kable(head(dfFlights))
```

```{r agfd}

dfFlights %<>%
  gather("Los.Angeles", "Phoenix", "San.Diego", "San.Francisco", "Seattle", key="City", value="NumOfFlights") %<>%
  fill(Airline)
```



<span style="color:darkorchid"><font size="6">__Principle 3. Less is more __ </font size></span>

## Why a box?

Consider this box plot.  What dos the box actually get you?  What is the ratio of data expression to ink?


```{marginfigure, echo=TRUE}
This boxplot is great for analysis ...
```

```{r dfs}
dfSummaryStats <- dfFlights %>% 
  group_by(Airline, City) %>% 
  summarize(sum_DelayedFlights=sum(NumOfFlights[Status=="delayed"]), sum_Flights = sum(NumOfFlights), propOfDelay=sum(NumOfFlights[Status=="delayed"])/sum(NumOfFlights)*100)

dfSummaryStats2 <- dfFlights %>% 
  group_by(Airline) %>% 
  summarize(sum_DelayedFlights=sum(NumOfFlights[Status=="delayed"]), sum_Flights = sum(NumOfFlights), propOfDelay=sum(NumOfFlights[Status=="delayed"])/sum(NumOfFlights)*100)

dfDelaysOnly <- dfFlights %>%
  filter(Status=="delayed")

dfDelaysOnly_Amwest <- dfDelaysOnly %>%
  filter(Airline=="AMWEST")

dfDelaysOnly_Alaska <- dfDelaysOnly %>%
  filter(Status=="ALASKA[")

dfPercentDiff <- dfSummaryStats %>%
  select(Airline, City, propOfDelay) %>%
  group_by(City) %>%
  summarize(propDiff=propOfDelay[Airline=="AMWEST"] - propOfDelay[Airline=="ALASKA"])

```

```{r boxplot}

gg2 <-ggplot(dfSummaryStats, aes(reorder(City, propOfDelay), propOfDelay)) +
geom_boxplot() + theme(axis.text.x = element_text(angle = 90), axis.title = element_text(size=15), axis.text = element_text(size = 15), plot.title = element_text(size=18)) +
  ggtitle("Percentage of Delayed Flights by City") +
  ylab("Percentage of Flights Delayed") +
  xlab("City") + 
  coord_flip()
gg2
```

```{marginfigure, echo=TRUE}

... but this is cleaner, leaner and perhaps clearer for a presentation.

```



```{r b2}

gg1 <- ggplot(dfSummaryStats, aes(reorder(City, propOfDelay), propOfDelay)) + theme_tufte() +
geom_tufteboxplot(outlier.colour="transparent") + theme(axis.title=element_blank(), axis.text.x = element_text(angle = 0), axis.text = element_text(size = 20), plot.title = element_text(size=20), axis.ticks=element_blank()) +
annotate("text", x = 0, y =30, adj=0,  family="Gill Sans MT", label = "") +
  ggtitle("Percentage of Delayed Flights by City") + 
  coord_flip()
gg1

```


__*Or consider this example:*__

```{r size}

g4 <- ggplot(dfSummaryStats, aes(x=reorder(City, propOfDelay), y=propOfDelay, group=Airline, fill=Airline)) +
  geom_col(position = position_dodge()) +
  ylab("City") + 
  xlab("Percentage of Delays") +
    ggtitle("Percent Delays by City and Airline") +
  coord_flip()
g4
```
      

   ... to this:
```{r xyz, fig.margin=TRUE, fig.cap="Fig.1: Percent delays by city and airline"}

ggplot(dfSummaryStats, aes(x=reorder(City, propOfDelay), y=propOfDelay, group=Airline, fill=Airline)) + theme_tufte(base_size=14, ticks=F) +
  geom_bar(position = position_dodge(), width=0.3, stat = "identity") +  theme(axis.title=element_blank()) +
  annotate("text", x = 3.5, y = 5, adj=1,  family="Gill Sans MT",
label = "") +
    theme(axis.text.x = element_text(angle = 0), axis.text = element_text(size = 24 , family="Gill Sans MT"), legend.text = element_text(size=20), legend.title=element_blank()) +
  coord_flip()

```

```{r ava}
g5 <- ggplot(dfSummaryStats, aes(x=reorder(City, propOfDelay), y=propOfDelay, group=Airline, fill=Airline)) + theme_tufte() +
  geom_bar(position = position_dodge(), width=0.3, stat = "identity") +  theme(axis.title=element_blank()) +
  annotate("text", x = 3.5, y = 5, adj=1,  family="Gill Sans MT",
label = "") +
    theme(axis.text.x = element_text(angle = 0), plot.caption = element_text(size=24), axis.text = element_text(size = 15 , family="Gill Sans MT")) +
      ggtitle("Percent Delays by City and Airline") +
  coord_flip()
g5
```

```{marginfigure, echo=TRUE}

<span style="color:darkorchid"><font size="6">__Principle 4. Graphs don't use language - they are language __ </font size></span>

```

```{marginfigure, echo=TRUE}
Consider these sparklines - they are graphical words.

```

```{r llg, fig.margin=TRUE}


library(reshape)
library(RCurl)
library(extrafont)
#font_import()
#loadfonts(device = "win")


dfFlightYears <- read_csv("D:\\RStudio\\CUNY_607\\Assigments\\Assignment 5 - Tufte\\FileForSlopeGraph.csv")
dfFlightYears %<>%
  gather("Alaska", "Amwest", "British", "Continental", "Delta", "Eastern", "Hawaiian", key="airline", value="flights")  %<>%
  mutate(flights=as.integer(flights))


dfFlightYears %<>%
  mutate(yearCat = as.character(as.integer(Year/10)*10))

dfFl <- dfFlightYears %>%
  group_by(airline, yearCat) %>%
  summarize(Flights=round(mean(flights)))

d <- dfFlightYears
mins <- group_by(d, airline) %>% slice(which.min(flights))
maxs <- group_by(d, airline) %>% slice(which.max(flights))
ends <- group_by(d, airline) %>% filter(Year == max(Year))
quarts <- d %>% group_by(airline) %>%
  summarize(quart1 = quantile(flights, 0.25),
            quart2 = quantile(flights, 0.75)) %>%
  right_join(d)

ggplot(d, aes(x=Year, y=flights)) + 
  facet_grid(airline ~ ., scales = "free_y") + 
  geom_line(size=0.7) +
  geom_text(data = ends, aes(label = airline), hjust = 0, nudge_x = 8, size=8, family="Gill Sans MT") +
  expand_limits(x = max(d$Year) + (0.7 * (max(d$Year) - min(d$Year)))) +
  scale_x_continuous(breaks = seq(1960, 2010, 10)) +
  scale_y_continuous(expand = c(0.2, 0)) +
  theme_tufte() +
  theme(axis.title=element_blank(), axis.text = element_blank(), 
        axis.ticks = element_blank(), strip.text = element_blank()) +
  theme(panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))
```

```{marginfigure, echo=TRUE}
In R we can put more data into a sparkline if we want to.  Here we see IQR and min/max graphed onto a sparkline:
```

```{r llout, fig.margin=TRUE}


library(reshape)
library(RCurl)
library(extrafont)
#font_import()
#loadfonts(device = "win")

d <- dfFlightYears
mins <- group_by(d, airline) %>% slice(which.min(flights))
maxs <- group_by(d, airline) %>% slice(which.max(flights))
ends <- group_by(d, airline) %>% filter(Year == max(Year))
begins <- group_by(d, airline) %>% filter(Year == min(Year))
quarts <- d %>% group_by(airline) %>%
  summarize(quart1 = quantile(flights, 0.25),
            quart2 = quantile(flights, 0.75)) %>%
  right_join(d)

ggplot(d, aes(x=Year, y=flights)) + 
  facet_grid(airline ~ ., scales = "free_y") + 
  geom_ribbon(data = quarts, aes(ymin = quart1, max = quart2), fill = 'grey90') +
  geom_line(size=0.7) +
  geom_point(data = mins, col = 'forestgreen', size=5) +
  geom_point(data = maxs, col = 'blue', size = 5) +
  geom_text(data = mins, aes(label = flights), vjust = -1, size=6, family="Gill Sans MT", col='forestgreen') +
  geom_text(data = maxs, aes(label = flights), vjust = 2.5, size=6, family="Gill Sans MT", col="blue") +
  #geom_text(data = ends, aes(label = flights), hjust = 0, nudge_x = 1, size=6, family="Gill Sans MT", col='red') +
  geom_text(data = ends, aes(label = airline), hjust = 0, nudge_x = 8, size=8, family="Gill Sans MT", col='black') +
  #geom_text(data = begins, aes(label = flights), hjust = 0, nudge_x = -1, size=6, family="Gill Sans MT", col='red') +
  expand_limits(x = max(d$Year) + (.5 * (max(d$Year) - min(d$Year)))) +
  scale_x_continuous(breaks = seq(1960, 2010, 10)) +
  scale_y_continuous(expand = c(.5, 0)) +
  theme_tufte() +
  theme(axis.title=element_blank(), axis.text.y = element_blank(), 
        axis.ticks = element_blank(), strip.text = element_blank(), axis.text = element_text(size = 14, family="Gill Sans MT")) +
  theme(panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))
```

Tufte invented the sparkline - a small graph that shows the upward ad downward trajectory of some variable in a small, concise graphical statement.  For Tufte, the sparkline functions just as a word does -  small, discrete representation meant to express a single - though complex - idea. 


<span style="color:darkorchid"><font size="6">__Principle 5. Don't live in a cage__ </font size></span>

In the graph below, the lines are the same as in a line graph.  But the y axis has been obliterated - instead, it is occupied by the names of airlines on either side with their traffic shown at various points. It is as if the sparkline words had been arranged in a paragraph. It is a very small change but a highly effective one:

```{r gv}


#install.packages('devtools')
#library(devtools)
#install_github("leeper/slopegraph")#install Leeper's package from Github
library(slopegraph)
ggslopegraph2(dataframe = dfFl,
                times = yearCat,
                measurement = Flights,
                grouping = airline,
                title = "Traffic by Airline, 1960 to 2010",
                caption = NULL,
                linecolor = "gray",
                subtitle = NULL, 
                linethickness = .5,
                ytextsize = 3
                )

```
